---
title: "ej_screen"
format: html
editor: visual
---

```{r}
# Loading in Libraries
library(tidyverse)
library(sf)
library(here)
library(tmap)
```

```{r}

# Read in geodatabase of EJScreen data at the Census Block Group level
ejscreen <- sf::st_read(here::here("data", "ejscreen","EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb")) 
```

```{r}
# Filter to only include California data
california <- ejscreen %>%
  dplyr::filter(ST_ABBREV == "CA") 

# Filter to only include Monterey County data
monterey <- ejscreen %>%
  dplyr::filter(CNTY_NAME %in% c("Monterey County"))
```

```{r}
# Turning PEOPCOLORPCT column into intergers from decimals
monterey$PEOPCOLORPCT <- monterey$PEOPCOLORPCT * 100
```


```{r}
# Generating Map 1
tm_shape(monterey) + #defines spacial object to plot
  tm_polygons() +
  tm_graticules() + # Adding graticules
  tm_fill("PEOPCOLORPCT",  # % unemployed
          palette = "Blues", 
          title = "POC Percentile (%)") + 
  tm_title("People of Color Percentile in Monterey County") +
  tm_scale_bar(position = c("left", "bottom")) +
  tm_compass(position = c("right", "top"))
```

```{r}
# Generating Map 2
tm_shape(monterey) + #defines spacial object to plot
  tm_polygons() +
  tm_graticules() + # Adding graticules
  tm_fill("P_RSEI_AIR",   # Percentile for Toxic Releases to Air
          palette = "Blues", 
          title = "Exposure Risk Percentile (%)") + 
  tm_title("Toxic Air Pollutant Exposure Risk in Monterey County") +
  tm_scale_bar(position = c("left", "bottom")) +
  tm_compass(position = c("right", "top"))
```
## About my Maps
The maps produced display different environmental injustices occurring in Monterey County, CA. Map 1 shows the percentile for people of color. The second map shows exposure risk for toxic air pollutants. When the maps are compared you can see that there tends to be more toxic air polutants in places that have a higher percentage of for people of color. Within Monterey County there are four point-source for air pollution According to the US EPA's Toxic Release Inventory (TRI) (Wade, 2022). These four TRI points are Dole Fresh Vegetables Inc, Fujifilm Ultra Pure Solutions Inc, Lhoist North America Natividad Plant, and US Army Garrison Fort Hunter Liggett Ranges. The locations of these four points are where air pollutant exposure risk is the highest.    







